
% First talk about supervisors
I want to start this section by thanking my main supervisor, Martin Hell.
During my PhD, he has always provided great guidance -- helpful, clear, and concise.
His door has always been open, and he has always been willing to provide help.
After our discussions, I have always left his office feeling encouraged, hopeful, and with a clear idea of how to proceed.
Not only does he possess the skills of a good supervisor, he is also a great friend and I have enjoyed having my office next door to him during my years at the university.

I also wish to thank my assistant supervisor Paul Stankovski Wagner for his help during my research, and for being another great office neighbour.
In addition to the perks of having friendly neighbours in general, having my office squeezed in between Martin's and Paul's offices ensured that I did not arrive \emph{too} late in the mornings -- in fear of their witty comments about my working hours.
I also want to express my thanks to my other assistant supervisor Ben Smeets for his valuable research input, in particular related to trusted computing.
Finally, I wish to thank Thomas Johansson for actually suggesting me to pursue a PhD.

During my time at the department, I have got to know many colleagues, particularly in the Crypto and Security group.
First of all I wish to thank Jonathan and Erik for all our geeky discussions:
Jonathan for our random technical discussions about everything from the strict aliasing rule to thick microcontroller manuals,
and Erik for his random math-related monologues in my office, which I have to admit I enjoy, even though he after several years still involuntarily sabotages my office door every time he visits.
I also wish to thank Pegah for giving me a worthy competitor in bringing home-baked Thursday fika, Alexander for interesting server management discussions, Nicolae for our research collaboration both at RISE and the university, and finally Carl for significantly increasing my vitamin~D production by forcing me to eat lunch outside.
During my time as a PhD student, the research group has grown significantly, so instead of trying to name all current and past members, I just wish to say that it has been a pleasure to get to know all of you, and sharing a great environment for research and small-talk.

% tech. and admin. staff
I also want to thank the administrative and technical staff at the department for helping me during my time at the department.
A special thanks to Erik Jonsson for interesting discussions, and patiently handling all my requests for opening firewall ports, leaving the department's firewall with more holes than a Swiss cheese.

% Finally friends and family
Finally, I also wish to thank my friends and family, you have all helped and supported me during my PhD studies.
In particular, I want to mention all inspiring code evenings with Alex, weekly Thursday lunches with Paul, and daily Slack competitions with Henrik.
To my mom Carina, and my dad Håkan: thank you for all your love and support throughout my life, and for always being there when I need you.

% Signature.
\begin{flushright}
	\emph{Linus}
	
	Lund, September 2019
\end{flushright}
