The following papers are included in this dissertation:

\begin{description}
  \item[Paper~I]
    \paperIref
  \item[Paper~II]
    \paperIIref
  \item[Paper~III]
    \paperIIIref
  \item[Paper~IV]
    \paperIVref
    
    This dissertation contains the full version of this paper, with extended descriptions and motivations of the recommender and its parameters.
  \item[Paper~V]
    \paperVref
  \item[Paper~VI]
    \paperVIref
\end{description}

\newpage
\noindent
The table below summarizes the responsibilities Linus Karlsson had in each paper:

\vspace{1em}
\begin{center}
\begin{tabular}{lllll}
  \toprule
  \emph{Paper} & \emph{Writing} & \emph{Concepts} &  \emph{Implementation} & \emph{Evaluation} \\
  \midrule
  \textbf{I}    & yes     & yes     & YES & -- \\ 
  \textbf{II}   & YES     & YES     & YES & -- \\
  \textbf{III}  & yes     & partial & yes & yes \\
  \textbf{IV}   & YES     & yes     & YES & yes \\
  \textbf{V}    & yes     & yes     & YES & YES \\
  \textbf{VI}   & YES     & partial & YES & YES \\
  \bottomrule
\end{tabular}
\end{center}
\vspace{1em}

\noindent
Capital letters indicate roles where Linus Karlsson took primary responsibility for the given role.
The individual contributions of Linus are described in more detail in the following paragraphs.

In Paper~I, Linus was involved in writing and concept design.
Linus solely constructed the implementation to test the proposed concept.

In Paper~II, Linus had the main responsibility for both writing and designing the proposed solution.
He also solely constructed the implementation to verify the solution.

In Paper~III, Linus was responsible for writing the sections about trust anchors in general, and everything related to the application plane.
% Sections~\ref{subsec:trust-anchors}, \ref{subsec:application-plane-trust}, \ref{subsec:trustanchors:appplane}, \ref{subsec:application-evaluation}.
He was partially responsible for the concepts in the paper: those related to the application plane.
Linus was also responsible for the implementation and performance evaluation of the application plane network element enrollment.

In Paper~IV, Linus was responsible for writing all of the system model and implementation sections.
Together with the other authors, he defined the overall goals of the recommender, and was then responsible for the detailed design of the system.
He was solely responsible for the implementation of the system.

In Paper~V, Linus was responsible for writing the sections about privacy profiles, isolated execution, implementation, and evaluation.
Together with the other authors, he defined the proposed privacy-preserving solution.
He was solely responsible for the implementation and evaluation of the solution.

In Paper~VI, Linus was the main responsible for writing of the complete paper.
He was partially involved in the design of the proposed algorithm.
Linus was solely responsible for the implementation of the newly proposed algorithms, as well as the evaluation of the results.

A further description of the papers' contributions \emph{to the research field} is presented in Section~\ref{sec:kappa-contributions}.

\newpage
\subsection*{Other Contributions}

The following peer-reviewed publications have also been published during my PhD studies, but are not included in this dissertation.

\begin{itemize}
	\item Christopher Jämthagen, Linus Karlsson, Paul Stankovski, and Martin Hell: ``eavesROP: Listening for ROP Payloads in Data Streams''. In \emph{Information Security Conference, ISC 2014, Hong Kong}, LNCS Vol. 8783, pp. 413--424, Springer.
	\item Linus Karlsson, Martin Hell, and Paul Stankovski. ``Improved Greedy Nonrandomness Detectors for Stream Ciphers''. In \emph{3rd International Conference on Information Systems Security and Privacy, ICISSP 2017, Porto, Portugal}. pp. 225--232, SCITEPRESS.
	\item Nicolae Paladi and Linus Karlsson. ``Safeguarding VNF Credentials with Intel SGX''. In \emph{SIGCOMM Posters and Demos '17, Los Angeles, CA, USA}. ACM.
	\item Alexander Cobleigh, Martin Hell, Linus Karlsson, Oscar Reimer, Jonathan Sönnerup, and Daniel Wisenhoff. ``Identifying, Prioritizing and Evaluating Vulnerabilities in Third Party Code''. In \emph{2018 IEEE 22nd International Enterprise Distributed Object Computing Workshop (EDOCW), Stockholm, Sweden}. pp. 208-211, IEEE.
\end{itemize}


%N. Paladi, L. Karlsson, K. Elbashir: “Trust Anchors in Software Defined Networks”. ESORICS 2018, Barcelona, Spain, LNCS Vol. 11009, pp. 485–504, Springer. DOI: 10.1007/978-3-319-98989-1_24
%[PDF]
%
%
%L. Karlsson, M. Hell, P. Stankovski: “Not So Greedy: Enhanced Subset Exploration for Nonrandomness Detectors”. Information Systems Security and Privacy. ICISSP 2017, Revised Selected Papers, CCIS Vol. 867, pp. 273–294, Springer. DOI: 10.1007/978-3-319-93354-2_13
%[PDF]
%
%L. Karlsson, M. Hell: “Enabling Key Migration Between Non-Compatible TPM Versions”. Trust and Trustworthy Computing, TRUST 2016, Vienna, Austria, LNCS Vol. 9824, pp. 101–118, Springer. DOI: 10.1007/978-3-319-45572-3_6
%[PDF]
%
%M. Hell, L. Karlsson, B. Smeets, J. Mirosavljevic: “Using TPM Secure Storage in Trusted High Availability Systems”. The 6th International Conference on Trusted Systems, INTRUST 2014, Beijing, China, LNCS Vol. 9473, pp. 243–258, Springer. DOI: 10.1007/978-3-319-27998-5_16
%[PDF]
%



%L. Karlsson, M. Hell, P. Stankovski: “Improved Greedy Nonrandomness Detectors for Stream Ciphers”. 3rd International Conference on Information Systems Security and Privacy, ICISSP 2017, Porto, Portugal, pp. 225–232, SCITEPRESS. DOI: 10.5220/0006268202250232
%[PDF]
%
%N. Paladi, L. Karlsson: “Safeguarding VNF Credentials with Intel SGX”. SIGCOMM Posters and Demos ’17, Los Angeles, CA, USA, ACM. DOI: 10.1145/3123878.3132016
%
%A. Cobleigh, M. Hell, L. Karlsson, O. Reimer, J. Sönnerup, D. Wisenhoff: “Identifying, Prioritizing and Evaluating Vulnerabilities in Third Party Code”. IEEE EDOCW 2018, Stockholm, Sweden, pp. 208–211, IEEE. DOI: 10.1109/EDOCW.2018.00038




% C. Jämthagen, L. Karlsson, P. Stankovski, M. Hell: “eavesROP: Listening for ROP Payloads in Data Streams”. Information Security Conference, ISC 2014, Hong Kong, LNCS Vol. 8783, pp. 413–424, Springer. DOI: 10.1007/978-3-319-13257-0_25
% [PDF]